---
title: "Worst Visualization Challenge"
author: "Amber Camp" 
date: "`r Sys.Date()`"
format: html
editor: visual
toc: true
---

## Welcome to the Worst Visualization Challenge!

Today, your job is to create a **terrible**, **awful** visualization.

Your plot should use real data but it should be **bad** in terms of communication, clarity, or aesthetics.

This is your chance to break all the rules… and then reflect on why those rules exist.

Things to (ab)use:

-   Bad color choices (neon on neon, anyone?)
-   Confusing labels
-   Misleading chart types (pie charts for time series, bar charts for continuous variables, etc.)
-   Overplotting and clutter
-   Illegible fonts or sizes
-   Nonsense axis scaling

Check online for resources on how to make *good* visuals, but here are a few really useful ones:

-   <https://ggplot2-book.org>
-   <https://r-graphics.org>
-   <https://r-graph-gallery.com>

## Load Libraries

We'll start with the basics. If you want to do more (like make a map), you can load relevant packages later

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(palmerpenguins)
```

## Today's Data Buffet

Let's bring in an assortment of familiar datasets: `mtcars`, `diamonds`, `penguins`, and `mpg`. Add more later, if you want (try `airquality` or `starwars`). Feel free to explore and use what inspires your worst work.

```{r, echo=FALSE}
data("mtcars")
data("diamonds")
data("penguins")
data("mpg")

head(mtcars)
head(diamonds)
head(penguins)
head(penguins_raw)
head(mpg)
```

## Demo Time

What do you expect to see below? How can we make this better?

```{r,warning=FALSE}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point() +
  labs(x = "Flipper Length",
       y = "Body Mass",
       title = "TITLE") +
  theme_minimal()

```

How about this one?

```{r, warning=FALSE}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 6, shape = 8, alpha = 0.75) +
  labs(title = "WHO EVEN KNOWS", x = "data", y = "more stuff") +
  theme(plot.title = element_text(family = "Comic Sans MS", color = "orange", size = 28),
        axis.text.x = element_text(angle = 145, color = "red"),
        panel.background = element_rect(fill = "hotpink"))
```

## What Counts As "Bad" (but good for today)?

### Allowed:

-   Confusing or clashing color schemes (love to see it)
-   Bad font choices
-   Too many slices/bars/points
-   Inappropriate chart types (bar chart of continuous data, etc.)
-   Bad axis labels or weird ordering

### Not Allowed:

-   Fake or renamed variables (e.g., don’t label penguins as "cats")
-   Non-working code

## Layers

The order of your layers matters. Use this to your advantage.

```{r, warning=FALSE}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=class)) + 
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") +
  geom_smooth(method="lm", se=FALSE, color="black") 

# you can reorganize the layers
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_smooth(method="lm", se=FALSE, color="black") +
  geom_point(aes(color=class)) + 
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") 




## note: geom_point() (etc) is actually a shortcut for the following:
# layer(mapping = NULL, 
#   data = NULL,
#   geom = "point", 
#   stat = "identity",
#   position = "identity")
```

You can also add layers to a "finished" plot:

```{r}
car_class <- ggplot(mpg) + 
  geom_bar(
    mapping = aes(x = manufacturer, fill = class),
    width = 1) + 
  theme(aspect.ratio = 1) +
  scale_fill_brewer(palette = "Set3")

car_class
car_class + coord_flip()
car_class + coord_polar()

car_models <- ggplot(mpg) + 
  geom_bar(
    mapping = aes(x = fct_infreq(manufacturer), fill = model), 
    show.legend = FALSE,
    width = 1) + 
  theme(aspect.ratio = 1) 

car_models
car_models + coord_flip()
car_models + coord_polar()
```

### Using `ggsave`

```{r, message=FALSE}
# ggsave("car_models.png", plot = car_models, width = 6, height = 4, dpi = 300)
ggsave("car_models.png", plot = car_models)
```

## Themes

```{r, message=FALSE}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_smooth(method="lm", se=FALSE, color="black") +
  geom_point(aes(color=class)) + 
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") +
  theme_dark()
```

### ggthemes

```{r, message=FALSE}
install.packages("ggthemes")
library(ggthemes)

ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=class)) + 
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") +
  geom_smooth(method="lm", se=FALSE, color="black") +
  theme_void()
```

### Custom Themes

```{r, warning=FALSE}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=class)) +
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") +
  theme(panel.background = element_rect(fill = "azure2"),
  plot.background = element_rect(fill = "white"),
  axis.text = element_text(size = 12, color = "slateblue", face = "italic"),
  axis.title = element_text(size = 14, face = "bold"),
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  panel.grid.major = element_line(color = "grey", size = 0.5),
  panel.grid.minor = element_line(color = "lightgrey", size = 0.25))

custom_theme <- theme(
  panel.background = element_rect(fill = "azure2"),
  plot.background = element_rect(fill = "white"),
  axis.text = element_text(size = 12, color = "blue", face = "italic"),
  axis.title = element_text(size = 14, face = "bold"),
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  panel.grid.major = element_line(color = "grey", size = 0.5),
  panel.grid.minor = element_line(color = "lightgrey", size = 0.25))

ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_boxplot(aes(color=class)) +
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") +
  custom_theme
```

### `extrafont`

```{r, message=FALSE}
install.packages("extrafont")
library(extrafont)

# very slow, but you only need to run below ONCE!
# font_import()
# loadfonts(device = "pdf")

ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=class), size=25) +
  labs(title="Engine Displacement vs. Highway MPG", x="Displacement (L)", y="Highway MPG") +
  theme_minimal() +
  theme(plot.title = element_text(family = "Comic Sans MS", size = 18, color = "burlywood", face = "bold", hjust = 0.5),
    axis.title = element_text(family = "Times New Roman", size = 14, face = "italic"),
    axis.text = element_text(family = "Courier New", size = 12, color = "red", face = "bold"),
    legend.title = element_text(family = "Bauhaus 93", size = 16, hjust = 0.5),
    legend.text = element_text(family = "Zapfino", size = 12),
    panel.background = element_rect(fill = "blueviolet"))
```

## My final gift: Pie Charts

There is no built in ggplot pie chart, but you can use `coord_polar()`!

(There are other ways: <https://r-graphics.org/RECIPE-MISCGRAPH-PIE.html>)

```{r}
mpg_models <- mpg %>%
  count(model)

ggplot(mpg_models, aes(x = "", y = n, fill = model)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y", start = pi/3) +
  scale_fill_manual(values = rainbow(nrow(mpg_models))) +
  geom_text(aes(label = model), position = position_stack(vjust = 0.5), size = 2, angle = 45) + 
  labs(title = "Chaos Pie", x = "???", y = "???") +
  theme(
    plot.title = element_text(family = "Papyrus", size = 22, color = "hotpink", face = "bold"),
    axis.title = element_text(family = "Times New Roman", size = 12, face = "italic"),
    axis.text = element_text(family = "Courier New", size = 8, color = "red", face = "bold"),
    legend.title = element_text(family = "Comic Sans MS", size = 16, color = "blue"),
    legend.text = element_text(family = "Zapfino", size = 6, color = "yellow"),
    panel.background = element_rect(fill = "limegreen"))



set.seed(55)
cols <- sample(rainbow(nrow(mpg_models)))

ggplot(mpg_models, aes(x = "", y = n, fill = model)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y", start = pi/3) +
  scale_fill_manual(values = cols) +
  geom_text(aes(label = model), position = position_stack(vjust = 0.1), size = 2, angle = 45) +
  labs(title = "Chaos Pie", x = "???", y = "???") +
  theme(
    plot.title = element_text(family = "Papyrus", size = 22, color = "hotpink", face = "bold"),
    axis.title = element_text(family = "Times New Roman", size = 12, face = "italic"),
    axis.text = element_text(family = "Courier New", size = 8, color = "red", face = "bold"),
    legend.title = element_text(family = "Comic Sans MS", size = 16, color = "blue"),
    legend.text = element_text(family = "Zapfino", size = 6, color = "yellow"),
    panel.background = element_rect(fill = "limegreen"))
```

## Your Turn!

Your turn to build terrible plots. Be ready to share one final, creatively awful plot at the end of class (and, most importantly, to understand and explain why it's bad!!)

Recommendation: Everyone in your group should be involved, but consider assigning primary roles like designer, coder, or presenter.

```{r}



```

## Ok, one last gift:

<https://viz.wtf> (old, but still fun!)
